---
title: "Alternatives and Added Value of rsonar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alternatives and Added Value of rsonar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The existing ecosystem

R code quality is addressed by many tools. Here is an honest overview of the existing landscape and where `rsonar` fits in.

---

## Underlying tools used by rsonar

### lintr — static analysis

`lintr` ([CRAN](https://cran.r-project.org/package=lintr)) is the reference R linter, maintained by the r-lib community. It detects:

- Style errors (naming, indentation, line length)
- Anti-patterns (`T` instead of `TRUE`, `1:length(x)` instead of `seq_along(x)`)
- Excessive cyclomatic complexity
- Dead or commented-out code

**Limitations in standalone usage:** lintr produces results in the console or text files, without aggregation, visual reporting, or debt calculation.

### styler — automatic formatting

`styler` ([CRAN](https://cran.r-project.org/package=styler)) applies the *tidyverse style guide*. It can reformat code automatically (`style_dir()`) or check compliance without modification (`dry = "on"`).

**Limitations in standalone usage:** no consolidated reporting, no native CI integration.

### covr — test coverage

`covr` ([CRAN](https://cran.r-project.org/package=covr)) measures test coverage line by line. It exports to Codecov, Coveralls and Cobertura (GitLab CI format).

**Limitations in standalone usage:** raw numeric results, no configurable Quality Gate threshold.

### goodpractice — packaging best practices

`goodpractice` ([GitHub](https://github.com/MangoTheCat/goodpractice)) checks ~25 R best practices: unused dependencies, deprecated functions, `T_and_F_symbol`, etc.

**Limitations in standalone usage:** package-oriented only, does not work on plain scripts.

---

## Alternatives to rsonar in the R ecosystem

### r-lib/actions — GitHub Actions for R

[r-lib/actions](https://github.com/r-lib/actions) offers pre-built GitHub actions (`check-r-package`, `test-coverage`, `lint`). Excellent GitHub integration, but:

- **GitLab not natively supported**
- No consolidated HTML report
- No debt calculation
- No configurable Quality Gate

### pak + rcmdcheck

`rcmdcheck` runs `R CMD check` and captures results. Very comprehensive for CRAN package validation, but:

- Does not handle style or technical debt
- Raw, non-visual results

### SonarQube Community Edition (native)

SonarQube supports R since version 9.x via the community plugin [sonar-r-plugin](https://github.com/Merck/thehivr). This plugin is:

- **Poorly maintained** (last significant activity 2021)
- Limited in rules (< 20 rules vs 100+ in lintr)
- Requires a dedicated SonarQube instance

### codemetar / codemeta

`codemetar` generates package metadata but performs **no quality analysis**.

---

## Comparison table

| Feature | rsonar | lintr alone | r-lib/actions | SonarQube +plugin |
|---|:---:|:---:|:---:|:---:|
| Static analysis | ✅ | ✅ | ✅ | ⚠️ limited |
| Style checking | ✅ | ⚠️ partial | ✅ | ❌ |
| Test coverage | ✅ | ❌ | ✅ | ✅ |
| Packaging best practices | ✅ | ❌ | ❌ | ❌ |
| **Interactive HTML report** | ✅ | ❌ | ❌ | ✅ |
| **Technical debt (SQALE)** | ✅ | ❌ | ❌ | ✅ |
| **Quality Gate** | ✅ | ❌ | ❌ | ✅ |
| **Analysis comparison** | ✅ | ❌ | ❌ | ✅ |
| **SARIF export** | ✅ | ❌ | ❌ | ❌ |
| **Trend tracking** | ✅ | ❌ | ❌ | ✅ |
| R scripts (non-package) | ✅ | ✅ | ❌ | ❌ |
| Native GitLab CI | ✅ | ⚠️ manual | ❌ | ✅ (server) |
| Native GitHub Actions | ✅ | ⚠️ manual | ✅ | ✅ (server) |
| **SonarQube JSON export** | ✅ | ❌ | ❌ | N/A |
| JUnit XML export | ✅ | ❌ | ✅ | ✅ |
| **Zero server required** | ✅ | ✅ | ✅ | ❌ |
| Pure R language | ✅ | ✅ | ✅ | ❌ Java |

---

## Added value of rsonar

### 1. Single entry point

One function, `sonar_analyse()`, replaces 4 separate calls. The result is a coherent S3 object that can be passed to `sonar_report()`, `quality_gate()`, `export_junit()`, `export_sarif()`, etc.

### 2. Visual report without infrastructure

`sonar_report()` generates an HTML report **without a server**, without Docker, without a database. It opens in the browser and can be archived as a CI artifact.

### 3. Quantified technical debt

The **SQALE model** (Software Quality Assessment based on Lifecycle Expectations) is the same used by SonarQube:

$$\text{Rating} = f\left(\frac{\text{Debt (min)}}{\text{Base effort (min)}}\right)$$

Each category (lint errors, style, missing coverage…) contributes to a total debt in minutes, allowing you to **prioritize fixes**.

### 4. Quality Gate without SonarQube

```{r}
# Block CI if coverage drops below 80% or lint errors exist
quality_gate(res,
  coverage_min    = 80,
  lint_errors_max = 0,
  fail_on_error   = TRUE  # quit(status = 1) in CI
)
```

### 5. Coexistence with existing SonarQube

`export_sonar_json()` produces a file compatible with SonarQube's **Generic Issue Import**. Organizations with an existing SonarQube instance for Java/Python can inject R results **without any additional plugin**.

### 6. SARIF for GitHub Code Scanning

`export_sarif()` outputs the industry-standard SARIF format, which integrates directly with GitHub Code Scanning, VS Code, and Azure DevOps.

### 7. Analysis comparison and trend tracking

`sonar_diff()` compares two analyses to detect regressions, while `sonar_trend()` persists metrics to a JSON file for historical tracking — just like SonarQube's project history.

### 8. Works on scripts, not just packages

Unlike `goodpractice` or `rcmdcheck`, `rsonar` analyzes any directory containing `.R` files, even without a `DESCRIPTION`.

---

## Projects that inspired rsonar

- **[lintr](https://github.com/r-lib/lintr)** — the tool itself, very comprehensive
- **[megalinter](https://github.com/oxsecurity/megalinter)** — orchestrates linters for 50+ languages in Docker
- **[sonar-scanner](https://github.com/SonarSource/sonar-scanner-cli)** — reference for the Quality Gate model
- **[qodana](https://www.jetbrains.com/qodana/)** — JetBrains code inspector, inspiration for the HTML report
- **[r-lib/actions](https://github.com/r-lib/actions)** — reference for R CI integration
