---
title: "GitLab & GitHub Auto-Fix with sonar_autofix()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GitLab & GitHub Auto-Fix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`sonar_autofix()` is a **generic function** for automatic code fixes with Git platform integration. It runs `sonar_fix()` to correct code issues, then commits changes and creates a Merge Request (GitLab) or Pull Request (GitHub).

The function auto-detects the CI environment:

| Platform | Detection | Output |
|----------|-----------|--------|
| GitLab | `GITLAB_CI` or `CI_PROJECT_ID` set | Merge Request via API v4 |
| GitHub | `GITHUB_ACTIONS` set | Pull Request via `gh` CLI |

## Quick start

### One-liner (auto-detect)

```{r}
library(rsonar)
sonar_autofix()
```

The function automatically detects whether you're in GitLab CI or GitHub Actions.

### GitLab CI

```{r}
sonar_autofix(provider = "gitlab")
```

Requires `GITLAB_TOKEN` or `CI_JOB_TOKEN` environment variable.

### GitHub Actions

```{r}
sonar_autofix(provider = "github")
```

Requires `gh` CLI installed and authenticated, or `GITHUB_TOKEN` set.

## CI pipeline examples

### GitLab CI

```yaml
rsonar-autofix:
  stage: fix
  when: manual
  script:
    - R -q -e "library(rsonar); sonar_autofix()"
  artifacts:
    paths:
      - sonar-fixes.json
```

### GitHub Actions

```yaml
rsonar-autofix:
  runs-on: ubuntu-latest
  steps:
    - uses: actions/checkout@v4
    - uses: r-lib/actions/setup-r@v2
    - name: Auto-fix
      run: |
        R -q -e "remotes::install_github('ddotta/rsonar')"
        R -q -e "library(rsonar); sonar_autofix()"
```

## Parameters

### Provider selection

```{r}
# Auto-detect
sonar_autofix(provider = "auto")

# Force GitLab
sonar_autofix(provider = "gitlab")

# Force GitHub
sonar_autofix(provider = "github")
```

### Branch and target

```{r}
sonar_autofix(
  branch = "auto/style-fix",
  target_branch = "develop"
)
```

### Fix categories

```{r}
# Only formatting and spacing
sonar_autofix(fixes = c("styler", "spacing", "cleanup"))
```

### Dry-run mode

```{r}
# Simulate without modifying anything
sonar_autofix(dry_run = TRUE)
```

## Auto-detected CI variables

### GitLab

| Variable | Used for |
|----------|----------|
| `GITLAB_CI` | Detect GitLab environment |
| `CI_PROJECT_ID` | Project ID for API |
| `CI_SERVER_URL` | GitLab server URL |
| `CI_API_V4_URL` | API v4 endpoint |
| `CI_DEFAULT_BRANCH` | Target branch |

### GitHub

| Variable | Used for |
|----------|----------|
| `GITHUB_ACTIONS` | Detect GitHub environment |
| `GITHUB_BASE_REF` | Target branch |
| `GITHUB_TOKEN` | API authentication |
| `GITHUB_RUN_ID` | Included in PR description |

## Token authentication

### GitLab

Token resolution order:
1. `GITLAB_TOKEN` environment variable
2. `PERSONAL_ACCESS_TOKEN` environment variable
3. `CI_JOB_TOKEN` (built-in GitLab CI token)

Setting up a Project Access Token:
1. `Settings > Access Tokens > Add new token`
2. Token name: `rsonar-autofix`
3. Role: `Maintainer`
4. Scopes: `write_repository`, `api`
5. Add to `Settings > CI/CD > Variables` as `GITLAB_TOKEN` (masked)

### GitHub

Token resolution order:
1. `GITHUB_TOKEN` environment variable (Actions built-in)
2. `gh` CLI authentication (if installed)

## How it works

```
sonar_autofix(provider = "gitlab")
   │
   ├── 1. sonar_fix() → auto-correct code (16 fix categories)
   ├── 2. git checkout -b auto/rsonar-fix
   ├── 3. git add + git commit
   ├── 4. git push -f origin auto/rsonar-fix
   └── 5. GitLab API → POST /merge_requests
        or
        gh pr create → Pull Request
```

## Architecture

The function uses a common engine for Git operations and platform-specific connectors:

- `.run_git_operations()` — Common: checkout, commit, push
- `.create_gitlab_mr()` — GitLab: Merge Request via API v4
- `.create_github_pr()` — GitHub: Pull Request via `gh` CLI

## Report

A JSON report is generated at `sonar-fixes.json`:

```json
{
  "files": {
    "scanned": 42,
    "modified": 18
  },
  "fixes": {
    "styler": 205,
    "spacing": 81,
    "true_false": 9
  },
  "provider": "gitlab",
  "branch": "auto/rsonar-fix"
}
```

## See also

- [Auto-Fix with sonar_fix()](auto-fix.html) for detailed fix categories
- [CI/CD Integration](ci-integration.html) for general pipeline configuration
- [Introduction to rsonar](introduction.html) for a general overview