---
title: "Introduction to rsonar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rsonar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why rsonar?

Teams developing in **Python, Java or JavaScript** benefit from [SonarQube](https://www.sonarqube.org/): a centralized tool that analyzes code quality, measures test coverage, detects *code smells* and calculates **technical debt**. R developers, however, must juggle several disparate tools.

`rsonar` bridges this gap by orchestrating the best R tools available:

| Dimension | Underlying tool | SonarQube equivalent |
|---|---|---|
| Static analysis | `lintr` | Issues (bugs, code smells) |
| Code style | `styler` | Style violations |
| Test coverage | `covr` | Code coverage |
| Best practices | `goodpractice` | Maintainability checks |
| Technical debt | SQALE model | Technical debt |
| Quality Gate | `quality_gate()` | Quality Gate |

---

## Installation

```{r}
# From GitHub (recommended)
remotes::install_github("ddotta/rsonar")
```

---

## Typical Workflow

### 1. Analyze a project

```{r}
library(rsonar)

# Full analysis (from the root of an R package)
res <- sonar_analyse("path/to/my/package")
```

The `sonar_analyse()` function:

1. Lists all `.R` files, excluding cache directories (`renv`, `packrat`, etc.)
2. Runs `lintr::lint_dir()` using the project's `.lintr` file if present
3. Checks style with `styler::style_file(..., dry = "on")`
4. Measures coverage with `covr::package_coverage()`
5. Runs `goodpractice::gp()` on the package
6. Computes **technical debt** using the SQALE model

### 2. View the console summary

```{r}
print(res)
```

### 2bis. Get a quick quality percentage in your IDE

For rapid local feedback during development (without CI):

```{r}
quality_score(".")
```

By default, `quality_score()` runs a fast analysis (coverage and goodpractice
disabled) and displays a percentage score plus the SQALE rating.

```
── rsonar — Quality Report ──────────────────────────────────
ℹ Project  : /home/user/mypackage
ℹ Analysis : 2026-04-20 14:32
ℹ Files    : 8 R file(s)

── Metrics ──────────────────────────────────────────────────
  Lint        : 3 issue(s) (1 err / 2 warn / 0 style)
  Style       : 1 non-compliant file(s)
  Coverage    : 72.4%
  Goodpractice: 2 failure(s)

── Technical Debt ───────────────────────────────────────────
🟡 SQALE rating: C
⏱  Estimated duration: 1.42h (85 min)
```

### 3. Generate the HTML report

```{r}
sonar_report(res, output = "quality.html")
```

The HTML report contains:

- **Dashboard** with consolidated metrics and SQALE A→E rating
- **Lint issues list** with severity, file, line and rule
- **Improperly formatted files** detected by styler
- **Technical debt breakdown** by category and in minutes

### 4. Check the Quality Gate

```{r}
gate <- quality_gate(res,
  coverage_min     = 80,   # 80% minimum coverage
  lint_errors_max  = 0,    # 0 errors tolerated
  style_issues_max = 0,    # all code must be formatted
  rating_min       = "C"   # minimum SQALE rating
)
print(gate)
```

```
── Quality Gate: PASSED ──────────────────────────────────────
✔ Coverage >= 80% [82.1%]
✔ Lint errors <= 0 [0]
✔ Style issues <= 0 [0]
✔ SQALE rating >= C [B]
```

### 5. Set up the project

```{r}
# Add the recommended rsonar .lintr
use_rsonar_lintr()

# Add a pre-configured GitLab CI pipeline
use_rsonar_ci("gitlab")

# Or GitHub Actions
use_rsonar_ci("github")
```

---

## Analyzing non-package projects (plain R scripts)

`rsonar` also works on projects that are **not** R packages:

```{r}
res <- sonar_analyse(
  "path/to/project",
  include_coverage     = FALSE,  # no tests/
  include_goodpractice = FALSE   # no DESCRIPTION
)
sonar_report(res)
```

---

## Comparing analyses

Use `sonar_diff()` to compare two analyses and detect regressions:

```{r}
baseline <- sonar_analyse(".", include_coverage = FALSE)
# ... make changes ...
current  <- sonar_analyse(".", include_coverage = FALSE)
diff     <- sonar_diff(current, baseline)
print(diff)
```

---

## Tracking trends over time

Use `sonar_trend()` in your CI pipeline to build a history of quality metrics:
```{r}
res <- sonar_analyse(".")
sonar_trend(res, file = "rsonar-history.json")
```

---

## Export for SonarQube

If your organization already uses SonarQube for other languages, you can inject `rsonar` results via the **Generic Issue Import**:

```{r}
export_sonar_json(res, "sonar-issues.json")
```

Then in `sonar-project.properties`:

```
sonar.externalIssuesReportPaths=sonar-issues.json
```

## Export for GitHub Code Scanning (SARIF)

```{r}
export_sarif(res, "results.sarif")
```

For GitLab CI with JUnit artifacts:

```{r}
export_junit(res, "junit-results.xml")
```

---

## Next steps

- [Alternatives and added value of rsonar](alternatives.html)
- [Complete CI/CD integration](ci-integration.html)
- [Understanding technical debt](technical-debt.html)
