---
title: "Understanding and Reducing Technical Debt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding and Reducing Technical Debt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What is technical debt?

**Technical debt** is the cost of additional work caused by shortcuts in the code: quick fixes, untested code, poor naming, unnecessary complexity. Like financial debt, it accumulates interest if not repaid.

`rsonar` quantifies this debt in **estimated remediation minutes**, using the **SQALE** (Software Quality Assessment based on Lifecycle Expectations) model used by SonarQube.

---

## The SQALE Model in rsonar

### Debt calculation

The total debt is the sum of costs per category:

$$D_{total} = D_{lint\_err} + D_{lint\_warn} + D_{style} + D_{coverage} + D_{gp}$$

With default costs:

| Category | Default unit cost |
|---|---|
| Lint error (`error`) | 30 min |
| Lint warning (`warning`) | 10 min |
| Lint style violation | 2 min |
| Improperly formatted file (styler) | 5 min |
| Goodpractice failure | 20 min |
| Missing coverage point | 5 min/point |

### SQALE rating calculation

The rating is computed from the **debt-to-base-effort ratio**:

$$ratio = \frac{D_{total}}{N_{files} \times 30 \text{ min}}$$

| Ratio | Rating | Meaning |
|---|---|---|
| < 5% | **A** | Excellent quality |
| 5–10% | **B** | Good quality |
| 10–20% | **C** | Acceptable quality |
| 20–50% | **D** | Poor quality |
| > 50% | **E** | Critical |

---

## Example analysis

```{r}
library(rsonar)

res  <- sonar_analyse(".")
debt <- debt_index(res)
print(debt)
```

```
── rsonar Technical Debt ───────────────────────────────────
ℹ Estimated duration: 2.17h (130 min)
ℹ SQALE rating: C

── Breakdown by category ────────────────────────────────────
             category issues minutes
1      Lint (errors)      2      60
2     Lint (warnings)     4      40
3     Best practices      1      20
4          Coverage       2      10
```

---

## Customizing costs

Default costs can be adjusted to match your organization's context:

```{r}
# Stricter costs for a critical project
debt <- debt_index(res,
  cost_lint_error     = 60,   # 1h per error
  cost_lint_warning   = 15,
  cost_style          = 10,
  cost_gp             = 30,
  coverage_target     = 90,   # Target: 90%
  cost_coverage_point = 8
)
```

---

## Remediation strategy

### Prioritization: return on investment

Sort by `debt_minutes / number_of_corrections_needed` ratio:

```{r}
debt <- debt_index(res)

# Display categories with the most debt
breakdown <- debt$breakdown[order(-debt$breakdown$minutes), ]
print(breakdown)
```

### Progressive approach (legacy debt)

For an existing project with a lot of debt, adopt an incremental approach:

```{r}
# Week 1: fix lint errors (highest impact)
# Week 2: add tests to increase coverage
# Week 3: reformat the code (styler::style_dir)
# Week 4: address warnings and goodpractice

# Track debt evolution over time
debt_week_1  <- debt_index(res_before)$minutes
debt_week_4  <- debt_index(res_after)$minutes
reduction_pct <- round(100 * (1 - debt_week_4 / debt_week_1), 1)
cat("Debt reduction:", reduction_pct, "%\n")
```

---

## Comparing analyses with sonar_diff()

Use `sonar_diff()` to see exactly what changed between two analysis runs:

```{r}
baseline <- sonar_analyse(".")
# ... make improvements ...
current  <- sonar_analyse(".")
diff     <- sonar_diff(current, baseline)
print(diff)
```

This shows delta metrics, new issues introduced, and issues fixed.

---

## Trend tracking in CI

### Using sonar_trend()

`sonar_trend()` appends each analysis to a JSON history file, making it easy
to monitor quality evolution across builds:

```{r}
res <- sonar_analyse(".")
sonar_trend(res, file = "rsonar-history.json")
```

In your `.gitlab-ci.yml`:

```yaml
rsonar-report:
  stage: report
  when: always
  script:
    - |
      Rscript -e "
      library(rsonar)
      res <- sonar_analyse('.')
      sonar_trend(res)
      sonar_report(res, output='rsonar-report.html', open=FALSE)
      "
  artifacts:
    paths:
      - rsonar-report.html
      - rsonar-history.json
    expire_in: 4 weeks
```

### GitLab — rating as a metric

In your `.gitlab-ci.yml`, expose the rating as a metric:

```yaml
rsonar-metrics:
  script:
    - |
      Rscript -e "
      library(rsonar)
      res  <- sonar_analyse('.')
      debt <- debt_index(res)
      cat('SQALE rating:', debt\$rating, '\n')
      cat('Debt minutes:', debt\$minutes, '\n')
      # Fail if rating is D or E
      if (debt\$rating %in% c('D', 'E'))
        stop('Insufficient SQALE rating: ', debt\$rating)
      "
```

---

## Best practices for maintaining low debt

1. **Configure `.lintr`** from the start of the project with `use_rsonar_lintr()`
2. **Format systematically** before each commit: `styler::style_dir("R")`
3. **Write tests alongside code**, not after
4. **Block regressions** in CI with `quality_gate(fail_on_error = TRUE)`
5. **Don't accumulate commented code** — use Git history
6. **Name clearly** functions and variables — `object_length_linter(30)`

> The best way to manage technical debt is not to accumulate it.
